---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(lubridate)
library(ggplot2)
library(plyr)
library(dplyr)
```

Download the zipfile, unzip it and load the csv file

```{r warning=FALSE}
zipUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
dir.create('./data')
download.file(zipUrl, destfile = "./data/activity.zip", method="curl")
unzip("./data/activity.zip", exdir = "./data")
activityData <- read.csv("./data//activity.csv")
```

Make the date column actual dates and remove NA measurements.

```{r}
activityData$date <- ymd(activityData$date)
filteredData <- activityData[!is.na(activityData$steps),]
```

## What is mean total number of steps taken per day?

Sum and plot steps by day
```{r}
sums <- aggregate(filteredData$steps, by=list(Day=filteredData$date), FUN=sum)
plot(sums, type="h", ylab="Steps")
```


```{r echo=FALSE}
paste("Mean steps per day:", mean(sums$x, na.rm=TRUE))
paste("Median steps per day:", median(sums$x, na.rm=TRUE))
```

## What is the average daily activity pattern?
Aggregate by interval and take the average

```{r}
intvlSums <- aggregate(list(steps=filteredData$steps), by=list(interval=filteredData$interval), FUN=sum)
intvlSums$steps <- intvlSums$steps / length(unique(activityData$date))
plot(sums, type="l")
```

```{r echo=FALSE}
paste("Interval with most number of step is:", intvlSums[intvlSums$steps == max(intvlSums$steps), 1])
```

## Imputing missing values

```{r echo=FALSE}
paste("Number of rows missing data: ", sum(is.na(activityData$steps)))
```

Setting missing rows to the mean for that interval

```{r}
grouped <- group_by(activityData, interval)
withMean <- mutate(grouped, meanIntvl=mean(steps, na.rm=TRUE))
withMean$steps <- ifelse(is.na(withMean$steps), withMean$meanIntvl, withMean$steps)
```


Sum and plot by day
```{r}
sums <- aggregate(withMean$steps, by=list(Day=withMean$date), FUN=sum)
plot(sums, type="h", ylab="Steps")
```


```{r echo=FALSE}
paste("Mean steps per day:", mean(sums$x, na.rm=TRUE))
paste("Median steps per day:", median(sums$x, na.rm=TRUE))
```

Estimating missing data seems to have no effect on mean and median steps per day

## Are there differences in activity patterns between weekdays and weekends?

Calculate day type and plot steps by type

```{r}
dayType <- mutate(withMean, dayCat=ifelse(weekdays(date) == "Saturday" | weekdays(date) == "Sunday", "weekend", "weekday"))

weekdays <- dayType[dayType$dayCat=="weekday",]
weekends <- dayType[dayType$dayCat=="weekend",]

weekendSums <- aggregate(list(steps=weekends$steps), by=list(interval=weekends$interval), FUN=sum)
weekends$steps <- weekends$steps / length(unique(activityData$date))

weekdaySums <- aggregate(list(steps=weekdays$steps), by=list(interval=weekdays$interval), FUN=sum)
weekdays$steps <- weekdays$steps / length(unique(activityData$date))

par(mfrow=c(2,1))
plot(weekendSums, type="l", main="Weekends")
plot(weekdaySums, type="l", main="Weekdays")

```

